---
title             : "Characterizing North Carolina's Deaf/Hard-of-Hearing Infants and Toddlers: Predictors of Vocabulary, Diagnosis, and Intervention"
shorttitle        : "ELSSP"


author: 
  - name          : "Erin Campbell"
    affiliation   : "1"
    corresponding : yes    
    email         : "erin.e.campbell@duke.edu"
    address       : "Address here"
  - name          : "Elika Bergelson"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Duke University"


keywords          : "keywords"
wordcount         : "X"

bibliography      : ["ELSSP.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, cache = FALSE, include = FALSE, echo=FALSE}
library(papaja) 
library(tidyverse)
library(knitr)
library(kableExtra)
library(citr)
library(broom)
library(stargazer)
library(glue)
library(magick)
library(cowplot)
library(dotwhisker)
library(car)
knitr::opts_chunk$set(fig.width=6, fig.height=6, comment=F, message=F, hide = T, warning = F, echo = F)
```

```{r read-in-data, echo = F, results = 'hide', message = F, comment = F, fig.show='hide'}
source('SM_functions.R')
source('functions_and_df_prep.R')
```

# Introduction
In the United States, 1-2 children are born with hearing loss, per 1,000 births [@cdc2018]. This translates to 114,000 Deaf or Hard of Hearing (DHH) children born in the U.S. per year [@martin2019]. Of these 114,000, ~90% will be born to hearing parents [@mitchell2004], in a home where spoken language is likely the dominant communication method. Depending on the type and degree of hearing loss and whether the child uses amplification, spoken linguistic input will be partially or totally inaccessible. Some of these children will develop spoken language within the range of their hearing peers [@verhaert2008; @geers2017], but many will face persistent spoken language deficits [@eisenberg2007; @moeller2007; @luckner2010; @sarchet2014], which may later affect reading ability [@kyle2010] and academic achievement [@qi2012; @karchmer2003]. 

Despite many excellent studies examining language development in DHH children, there is still a gap in the literature describing and analyzing spoken language development across the full range of children receiving services for hearing loss, with many studies focusing in on specific subgroups (e.g. children under age X with Y level of hearing loss and Z amplification approach, e.g. @vohr2008; @yoshinaga-itano2018). In what follows, we first summarize the previous literature on predictors of spoken language outcomes in DHH children. We then provide  a brief overview of a common vocabulary measure used in the current study, the MacArthur-Bates Communicative Development Inventory (CDI). Finally, we turn to an empirical analysis of  early vocabulary in a wide range of young children receiving state services in North Carolina. We have two broad goals in what follows. First, we aim to provide a comprehensive description of a heterogeneous group of young children who receive state services for hearing loss. Second, we aim to connect the intervention approaches and child characteristics of this sample with children’s spoken vocabulary\footnote{Despite exciting, increasing, and converging evidence for benefits of early sign language exposure [e.g., @schick2007; @spencer1993; @davidson2014; @magnuson2000; @clark2016; @hrastinski2016], the majority of DHH children will not be raised in a sign language environment. This is particularly true for North Carolina, which does not have a large community of sign language users, relative to states like Maryland or areas like Washington D.C. or Rochester, NY. For this reason, and because no child in our sample used a legitimate signed language, we focus on spoken language development.}, with the broader goal of considering the success of early diagnosis and intervention initiatives. 

## Predictors of Language Outcomes
Though the literature points towards spoken language delays and deficits for DHH children, this is a highly variable population with highly variable outcomes [@pisoni2018]. Previous research indicates that gender [@kiese-himmel2002a; @ching2013], additional disability [@ching2013; @yoshinaga-itano2017; @verhaert2008], degree and configuration of hearing loss [@ching2013; @dediego-lazaro2018; @vohr2011; @yoshinaga-itano2017], amplification [@walker2015], communication [@geers2017], and early diagnosis/intervention [@yoshinaga-itano2017; @yoshinaga-itano2018] predict language outcomes in DHH children. We first provide a brief literature review on the effect of these predictors on language skills in DHH children.

### Gender
For hearing children, the literature points to a female gender advantage in early language acquisition. Girls speak their first word earlier [@macoby1966], have a larger [@bornstein2004; @fenson1994; @frank2017] and faster-growing vocabulary [@huttenlocher1991], and stronger grammatical and phonological skills [@lange2016; @ozcaliskan2010]. This finding appears to be consistent across studies [@wallentin2009], various spoken languages [@frank2019], and gesture [@ozcaliskan2010]. 

The DHH literature presents a more mixed (though rather understudied) picture. On one hand, DHH girls,  like hearing girls, have been found to have a larger spoken vocabulary than DHH boys [@kiese-himmel2002a; @ching2013]. However, in contrast to their hearing peers, DHH children do not seem to show a gender-based difference for some aspects of syntactic development [@pahlavannezhad2014].

### Comorbidities
Additional co-morbid disabilities occur frequently in the DHH population, perhaps as much as three times more than in the hearing population [@pollack1997]. Incidence estimates for co-occurring disabilities in DHH children range from 25-51% [@bruce2015; @guardino2008; @luckner2001; @picard2004; @schildroth1996; @soukup2007; @holden-pitt1998], with approximately 8% of DHH children living with 2 or more co-occurring disabilities [@schildroth1996]. 

Some of these conditions, particularly those which carry risk of developmental delay (e.g., Down syndrome), result in language delays independent of hearing loss [@chapman1997; @kristoffersen2008; @weismer2010]. These effects vary by the nature of the specific disability [@cupples2018; @cupples2014], with cognitive ability more predictive of language outcomes than presence or absence of additional disability [@meinzen-derr2011; @sarant2008]. Disability and hearing loss likely each contribute to a given child’s spoken language development [@rajput2003; @vannierop2016; @ching2013], with differential effects of each [@vesseur2016]. In some cases, additional disabilities appear to interact with hearing loss to intensify developmental delays [@birman2012; @pierson2007].

Furthermore, incidence of hearing loss is higher among children born premature (defined as < 37 weeks gestational age). Compared to an incidence of 0.2% in full-term infants, incidence of hearing loss in extremely premature infants (defined as < 33 weeks gestational age) ranges 2--11%, with increased prematurity associated with increased rates of hearing loss [@wroblewska-seniuk2017]. 

Independently of hearing status, prematurity is linked to increased risk of language delay and disorder [@rechia2016; @vannoort-vanderspek2012; @carter2017; @vohr2014; @cusson2003; @barre2011]. Unfortunately, research on language development in premature DHH children is scant [@vohr2016], so it remains unclear how hearing loss and prematurity may interact within spoken language skills. One study of premature infants finds that auditory brainstem response during newborn hearing screening predicts language performance on the PLS-4 at age 3 [@amin2014], suggesting a link between prematurity, hearing loss, and language development in early childhood, though further research is needed in this domain. In extremely premature DHH children, incidence of additional disabilities may be as high as 73% [@robertson2009]. Indeed, pre-term infants with comorbidities have been found to be more likely to also have hearing loss than those without comorbidities [@schmidt2003], further complicating language development for this population.

### Audiological Characteristics
Hearing loss varies in severity, ranging from slight to profound [@clark1981]. More severe hearing loss (less access to spoken language) typically results in more difficulty with spoken language in infancy [@vohr2008], early childhood [@ching2010; @ching2013; @tomblin2015; @sarant2008; @sininger2010] and school-age children [@wake2004]. Although profound hearing loss is associated with more pronounced spoken language difficulty, even mild to moderate hearing loss is associated with elevated risk of language disorders [@delage2007; @blair1985].

Hearing loss also varies in whether it affects one ear or both. Bilateral hearing assists speech perception, sound localization, and loudness perception in quiet and noisy environments [@ching2007]. The literature on hearing aids and cochlear implants also points to benefits for bilateral auditory input [@sarant2014; @lovett2010; @smulders2016]. At school-age, 3--6% of children have unilateral hearing loss [@ross2010]. Although children with unilateral hearing loss have one “good ear,” even mild unilateral hearing loss has been tied to higher risk of language delays and educational challenges relative to hearing children [@vila2015; @lieu2013; @lieu2004; @lieu2012; @kiese-himmel2002]. Just as in the bilateral case, more severe hearing loss leads to greater deficits in spoken language and educational outcomes for children with unilateral hearing loss [@lieu2013;@anne2017]. 

Many DHH children receive hearing aids (HAs) or cochlear implants (CIs) to boost access to the aural world. These devices have been associated with better speech perception and spoken language outcomes [@walker2015; @waltzman1997; @niparko2010]. In turn, aided audibility predicts lexical abilities with children in HAs [@stiles2012]. 

For both hearing aids and cochlear implants, earlier fit leads to better spoken language skills, if the amplification is effective. For hearing aids, some studies find that children with milder hearing loss who receive hearing aids earlier have better early language achievement than children who are fit with hearing aids later [@tomblin2015], but this finding does not hold for children with severe to profound hearing loss [@kiese-himmel2002; @watkin2007] (for whom hearing aids are generally ineffective). Analogously, children who are eligible and receive cochlear implants earlier have better speech perception and spoken language outcomes than those implanted later [@dettman2007; @miyamoto2008; @svirsky2004; @yoshinaga-itano2018; @artieres2009], with best outcomes for children receiving implants before their first birthday [@dettman2007]. 

### Communication
Total Communication refers to communication that combines speech, gesture, and elements of sign, sometimes simultaneously.  Total communication, while it often includes elements of sign, such as individual signs, is not a sign language, such as American Sign Language [@scott2020; @mueller2013]. Clinicians currently employ total communication as an alternative or augmentative communication method for children with a wide range of disabilities [@branson2009; @gibbs1991; @mirenda2003].

Compared to total communication, DHH children using an exclusively oral approach have better speech intelligibility [@geers2017; @dillon2004; @geers2002; @hodges1999] and auditory perception [@geers2017; @odonoghue2000]. That said, there is some debate as to whether an oral approach facilitates higher spoken language performance, or whether children who demonstrate aptitude for spoken language are steered towards the oral approach rather than total communication [@hall2017].

### 1-3-6 Guidelines
Early identification [@yoshinaga-itano2018; @yoshinaga-itano1998; @white1987; @robinshaw1995; @apuzzo1995; @kennedy2006] and timely enrollment in early intervention programs [@vohr2008; @watkin2007; @ching2013; @holzinger2011; @vohr2011; @ching2018] are associated with better language proficiency. Indeed, DHH children who receive prompt diagnosis and early access to services have been found to meet age-appropriate developmental outcomes, including language [@stika2015]. 

In line with these findings, the American Academy of Pediatricians (AAP) has set an initiative for Early Hearing Detection and Intervention (EHDI). Their EHDI guidelines recommend that DHH children are screened by 1 month old, diagnosed by 3 months old, and enter early intervention services by 6 months old. We refer to this guideline as 1-3-6. Meeting this standard appears to improve spoken language outcomes for children with HL [@yoshinaga-itano2017; @yoshinaga-itano2018] and the benefits appear consistent across a range of demographic characteristics. 

At a federal level in the U.S., the Early Hearing Detection and Intervention Act of 2010 [@capps2009] was passed to develop state-wide systems for screening, evaluation, diagnosis, and “appropriate education, audiological, medical interventions for children identified with hearing loss,” but policies for early diagnosis and intervention vary by state.  As of 2011, 36 states (including North Carolina, @2000] mandate universal newborn hearing screening [@nationalconferenceofstatelegislatures2011]. All states have some form of  early intervention programs that children with hearing loss can access [@nad], but these also vary state-by-state. For instance, half of the states in the US do not consider mild hearing loss an eligibility criterion for early intervention [@holstrum2008].

In evaluating the success of this initiative, the AAP [@ehdi] finds that about 70% of US children who fail their newborn hearing screening test are diagnosed with hearing loss before 3 months old, and that 67% of those diagnosed (46% of those that fail newborn hearing screening) begin early intervention services by 6 months old. These findings suggest that there may be breaks in the chain from screening to diagnosis and from diagnosis to intervention, and the effect may be further delays in language development for children not meeting these guidelines.
 
## Quantifying vocabulary growth in DHH children
The MacArthur Bates Communicative Development Inventory [CDI, @fenson1994] is a parent-report instrument that gathers information about children’s vocabulary development. The Words and Gestures version of the form is normed for 8--18-month-olds. On Words and Gestures, parents indicate whether their child understands and/or produces each of the 398 vocabulary items, and answer questions about young children’s early communicative milestones.  The Words and Sentences version of the form is normed for 16-30-month-olds. On Words and Sentences, parents indicate whether their child produces each of the 680 vocabulary items, and answer some questions about grammatical development. The CDI has been normed on a large set of participants across many languages [@anderson2002; @frank2017; @jackson-maldonado2003]. 

The CDI has also been validated for DHH children with cochlear implants [@thal2007]. More specifically, in this validation, researchers asked parents to complete the CDI, administered the Reynell Developmental Language Scales, and collected a spontaneous speech sample. All comparisons between the CDI and the other measures yielded significant correlations ranging from 0.58 to 0.93. Critically, the children in this study were above the normed age range for the CDI, and thus this validation helps to confirm that the CDI is a valid measurement tool for older DHH children. In further work, @castellanos2016 finds that in children with CIs, number of words produced on the CDI predicts language, executive function, and academic skills up to 16 years later. Building on this work, several studies have used the CDI to measure vocabulary development in DHH children (@yoshinaga-itano2017; @yoshinaga-itano2018; @dediego-lazaro2018; @vohr2008; @vohr2011; summarized in \@ref(tab:CDIsummary)). 


```{r CDIsummary}
research_table <- tribble(
  ~Study, ~Population,  ~Gender, ~'1-3-6', ~Laterality, ~Degree, ~Amplification, ~Communication, ~Comorbidities,
  "Yoshinaga-Itano et al., 2017", "8-39 month children with bilateral hearing loss", "No effect", "1-3-6 +", "Did not study", "More severe -", "Did not study", "Did not study", "Comorbidities -",
  "Yoshinaga-Itano et al., 2018", "Children with cochlear implants", "Did not study", "1-3-6 +", "Did not study", "Did not study", "Earlier CI activation +", "Did not study", "Did not study",
  "De Diego-Lazaro et al., 2018", "Spanish speaking children with bilateral hearing loss", "No effect", "Earlier intervention +", "Did not study", "Milder +", "More functional hearing +", "Did not study", "Did not study",
  "Vohr et al., 2011", "18-24 month olds with hearing loss", "Did not study", "Earlier intervention +", "Did not study", "Milder +", "Did not study", "Did not study", "NICU stay -; Comorbidities -"
)
kable(research_table,
      caption = "Summary of findings of CDI studies in DHH children") %>%
  kable_styling(latex_options = c("scale_down")) %>%
  add_footnote("+ equals bigger vocab, - equals smaller vocab")
```

## Goals and Predictions
This study aims to 1) characterize the demographic, audiological, and intervention variability in the population of DHH children receiving state services for hearing loss; 2) identify predictors of vocabulary delays; and 3) evaluate the success of early identification and intervention efforts at a state level. We include three subgroups of DHH children traditionally excluded from studies of language development: children with additional disabilities, children with unilateral hearing loss, and children from bilingual or non-English-speaking households [e.g., @yoshinaga-itano2018; @nicholas2006]. 

For the first goal, we had reason to expect that many of these variables would be related, due to known causal relations (e.g., cochlear implants recommended for severe hearing loss, but not mild hearing loss). We sought to provide descriptive documentation about the distrubution of demographic, audiological, and intervention characteristics in a diverse sample of DHH children receiving state services. For the second, we hypothesized that male gender, more severe degree of hearing loss,  bilateral hearing loss, no amplification use, prematurity, and presence of additional disabilities would predict larger spoken vocabulary delay. We did not have strong predictions regarding the effects of communication method or presence of other health issues (e.g., congenital heart malformation) on vocabulary. For the third goal, we hypothesized that children with less residual hearing (i.e., bilateral, more severe) and no co-occurring conditions would be earlier diagnosed and earlier to begin language services, and that earlier diagnosis would predict earlier intervention.

# Methods
``` {r demo-data}
num_subs <- nrow(elssp)
num_repeat_subs <- nrow(elssp %>% filter(as.numeric(VisitTotal) > 1)) 
num_girls <- nrow(elssp %>% filter(Gender == 'female'))
num_boys <- nrow(elssp %>% filter(Gender == 'male'))
min_age <- min(elssp$Age)
max_age <- max(elssp$Age)
mean_age <- mean(elssp$Age)
sd_age <- sd(elssp$Age)
num_too_old_wg <- nrow(elssp_curves %>% filter(CDIversion == 'WG' & Age>18))
num_spanish_cdi <- nrow(elssp %>% filter(PrimaryLanguage == 'Spanish'))
```

Clinical evaluations were obtained through an ongoing collaboration with  the North Carolina Early Language Sensory Support Program (ELSSP), an early intervention program serving children with sensory impairments from birth to 36 months. ELSSP passed along deidentified evaluations to our team after obtaining consent to do so from each family. No eligibility criteria beyond hearing loss and receiving an ELSSP evaluation were imposed, given our goal of characterizing the full range of DHH children with hearing loss in North Carolina.

The clinical evaluations included demographic and audiological information, CDI vocabulary scores, and the results of any clinical assessments administered (e.g., PPVT), detailed further below. For some children (n=`r num_repeat_subs`), multiple evaluations were available from different timepoints. In these cases, only the first evaluation was considered for this study, due to concerns regarding within-subjects variance for statistical analysis.

While this collaboration is ongoing, we opted to pause for this analysis upon receiving data from 100 children. Thus, the reported sample below consists of `r num_subs` children (`r num_boys` male / `r num_girls` female) ages `r min_age`--`r max_age` months (M=`r mean_age`, SD=`r sd_age`). Race and SES information were not available. Families were administered either the Words and Gestures or Words and Sentences version of the CDI based on clinician judgement. Children who were too old for Words and Gestures, but who were not producing many words at the time of assessment, were often given Words and Gestures (n=`r num_too_old_wg`). Families for whom Spanish was the primary language (n=`r num_spanish_cdi`) completed the Spanish language version of the CDI [@jackson-maldonado2003].

```{r CDIinfo, echo=FALSE}
tribble(
  ~'CDI version', ~'Average Age (SD)',  ~'Average Comprehension (SD)', ~'Average Production (SD)', ~'% Developmental Delays',
  glue('WG (n={nrow(elssp %>% filter(CDIversion=="WG"))})'), glue('{round(mean((elssp %>% filter(CDIversion=="WG"))$Age), digits=2)} ({round(sd((elssp %>% filter(CDIversion=="WG"))$Age), digits=2)}) months'), glue('{round(mean((elssp %>% filter(CDIversion=="WG"))$ComprehensionCDI, na.rm = TRUE), digits=0)} ({round(sd((elssp %>% filter(CDIversion=="WG"))$ComprehensionCDI, na.rm = TRUE), digits=1)}) words'), glue('{round(mean((elssp %>% filter(CDIversion=="WG"))$ProductionCDI, na.rm = TRUE), digits=0)} ({round(sd((elssp %>% filter(CDIversion=="WG"))$ProductionCDI, na.rm = TRUE),digits=1)}) words'), glue('{round(((nrow((elssp %>% filter(CDIversion=="WG")) %>% filter(DevelopmentalConcerns=="yes")))/(nrow((elssp %>% filter(CDIversion=="WG")))))*100, digits=2)}%'),
  glue('WS (n={nrow((elssp %>% filter(CDIversion=="WS")))})'), glue('{round(mean((elssp %>% filter(CDIversion=="WS"))$Age), digits=2)} ({round( sd((elssp %>% filter(CDIversion=="WS"))$Age), digits=2)}) months'), 'NA', glue('{round(mean((elssp %>% filter(CDIversion=="WS"))$ProductionCDI, na.rm = TRUE), digits=0)} ({round(sd((elssp %>% filter(CDIversion=="WS"))$ProductionCDI, na.rm = TRUE), digits=1)}) words'), glue('{round(((nrow((elssp %>% filter(CDIversion=="WS")) %>% filter(DevelopmentalConcerns=="yes")))/(nrow((elssp %>% filter(CDIversion=="WS")))))*100, digits=2)}%')) %>%
kable(caption = "CDI details") %>%
  kable_styling(latex_options = c("scale_down"))
```

With regard to comorbid diagnoses, children in this sample were coded as yes/no for cognitive development concerns (e.g., Down syndrome, global developmental delays; Cornelia de Lange syndrome), yes/no for prematurity (i.e., more than 3 weeks premature), yes/no for health issues (e.g., heart defects, kidney malformations, VACTERL association), and yes/no for vision loss (not corrected to normal by surgery or glasses). 

```{r comorbid-info}
perc_prem_NICU <- (nrow(comorbid%>%filter(NICU==1&IsPremature==1))/(nrow(elssp %>% filter(IsPremature==1))))*100
comorbid_table <- tribble(
  ~Condition, ~'Specific Condition',  ~n,
  "Premature", "", n_condition(IsPremature),
  "", "Extremely Premature", n_condition(extremelypremature),
  "", "NICU stay", n_condition(NICU),
  "Health Issues", "", n_condition(HealthIssues),
  "", "Heart", n_condition(heart_issues),
  "", "Lung", n_condition(lung_issues),
  "", "Illness", n_condition(illness),
  "", "Feeding Issues", n_condition(feeding_issues),
  "", "Pregnancy/Birth Complications", n_condition(preg_birth_comp),
  "", "Musculoskeletal", n_condition(mus_skel),
  "", "Cleft Lip/Palate", n_condition(cleft),
  "", "Other", n_condition(health_issues_other),
  "Developmental Concerns", "",  n_condition(DevelopmentalConcerns),
  "", "Down Syndrome", n_condition(downsyndrome),
  "", "Chromosomal Issues", n_condition(otherchromissues),
  "", "Neural Tube Defects", sum(n_condition(encephalocelecyst), n_condition(occipitalencephalocele)),
  "", "Other", sum(n_condition(agenesiscorpuscallosum), n_condition(oculomotorapraxia), n_condition(ventricular_issues), n_condition(corticalvisualimpairment), n_condition(hydrocephalus), n_condition(overalldevconcerns), n_condition(corneliadelange)),
  "Vision Loss", "",  n_condition(VisionLoss),
  "", "Retinopathy of Prematurity",  n_condition(RetinopathyOfPrematurity),
  "", "Nearsightedness",  n_condition(myopia),
  "", "Farsightedness",  n_condition(hyperopia),
  "", "Cortical Visual Impairment",  n_condition(corticalvisualimpairment)
)
kable(comorbid_table, longtable=T,
      caption = glue::glue("Additional Diagnoses (n={n_condition(anycomorbid)})")) %>%
  kable_styling(latex_options = "hold_position") %>%
  add_footnote("N.B. ns do not sum to total because many participants had multiple diagnoses")
```

Degree of hearing loss was most often reported with a written description (e.g., “mild sloping to moderate” or “profound high frequency loss”). We created 2 variables: hearing loss in the better ear, hearing loss in the worse ear, and average hearing loss (average of better and worse ear). For the analyses in this paper, we primarily use hearing loss in the worse ear to avoid any redundancies with laterality. Using the ASHA hearing loss guidelines, each of these was coded with a dB HL value corresponding with the median dB HL for the level of hearing loss (e.g., moderate hearing loss was coded as 48 dB HL), and sloping hearing loss was coded as the average of the levels (e.g. mild to moderate was coded as 40.5 dB HL). Participants were also coded for unilateral or bilateral hearing loss; presence or absence of Auditory Neuropathy Spectrum Disorder; and etiology of hearing loss (sensorineural, conductive, or mixed). Amplification was recorded as the device the child used at the time of assessment: either hearing aid, cochlear implant, or none.

```{r amp-info}
elssp %>% 
  filter(Amplification != '', Laterality != '') %>%
  dplyr::select(VIHI_ID, AgeImplantation, AgeAmplification, HLbetter, HLworse, Laterality, Amplification) %>%
  group_by(Laterality, Amplification) %>%
  dplyr::summarise(mean_HLbetter = round(mean(HLbetter, na.rm = TRUE), 2),
                   mean_HLworse = round(mean(HLworse, na.rm = TRUE), 2),
                   mean_age_amplification = round(mean(AgeAmplification, na.rm = TRUE), 2),
                   mean_age_implantation = round(mean(AgeImplantation, na.rm = TRUE), 2)) %>% 
  kable(caption = glue::glue("Audiological Characteristics of the Sample")) %>%
  kable_styling(latex_options = c("hold_position", "scale_down")) 
```

Communication method was recorded as spoken language, total communication, or cued speech. One participant had a parent fluent in sign language, but the reported communication method in the home was total communication. No child in our sample used American Sign Language or another signed language. The forms also listed the primary language spoken at home, which we binned into English-speaking and non-English-speaking. `r round((nrow(elssp%>%filter(PrimaryLanguage=="English"))/(nrow(elssp))*100))`% of families spoke English, and `r round((nrow(elssp%>%filter(PrimaryLanguage=="Spanish"))/(nrow(elssp))*100))`% spoke Spanish. For one child, who was adopted from another country after her second birthday, we recorded the language background as non-English-speaking, although the child's adoptive parents are English-speaking, because the child had lived most of her life in a non-English-speaking environment. 

```{r comm-info}
kable((language_stats <- elssp %>% 
  filter(PrimaryLanguage != '', Communication != '') %>%
  dplyr::select(VIHI_ID, Communication, PrimaryLanguage) %>%
  group_by(Communication, PrimaryLanguage) %>%
  dplyr::count(name = 'Tally') %>%
  pivot_wider(names_from = PrimaryLanguage, values_from = Tally) %>%
  replace_na(list(English=0, Hindi=0, Spanish=0)) %>%
  mutate(Total = English + Hindi + Spanish)),
      caption = "Language and communication characteristics of the sample") %>%
  kable_styling(latex_options = "hold_position")
```


Age at screening was measured as the child’s age in months at their first hearing screening. Age at screening was available for `r nrow(elssp %>% filter(NBHS=='Y'))` participants. All participants with a screening age available were screened at birth or while in the NICU. We presume that the vast majority of participants without age at screening received their screening as newborns, as North Carolina boasts a 98% NBHS rate [@ncdhhs2013]. Age at diagnosis was taken as the age in months when children received their first hearing loss diagnosis. All children were enrolled in birth-to-three early intervention services through ELSSP, and the date of enrollment was listed on the clinician evaluation. For determining whether participants met the 1-3-6 guidelines, given the very high rates of early screening reported in our sample and by the state, we imputed missing data by assuming that children met the "screening by 1 month" criterion if they met the "diagnoses by 3 months" and "service enrollment by  6 months" criteria. Finally, we also calculated the number of hours of early intervention services received per month (including service coordination, speech therapy, and occupational therapy, among others) based on the clinician report.

```{r meets136-info}
kable(tribble(
  ~'', ~'', 
'Diagnosis by 3 months', glue('{round((nrow(elssp%>%filter(meets13=="yes"))/(nrow(elssp %>% filter(meets13=="yes"|meets13=="no"))))*100,2)}%'),
'Average Age Diagnosis (SD)', glue('{round(mean(elssp$IdentificationOfHLMonths, na.rm = TRUE), digits=2)} ({round(sd(elssp$IdentificationOfHLMonths, na.rm = TRUE), digits=2)}) months'),
'Intervention by 6 months', glue('{round((nrow(elssp%>%filter(meets6=="yes"))/(nrow(elssp %>% filter(meets6=="yes"|meets6=="no"))))*100,2)}%'),
'Average Age Intervention (SD)', glue('{round(mean(elssp$AgeStartedServices, na.rm = TRUE), digits=2)} ({round(sd(elssp$AgeStartedServices, na.rm = TRUE), digits=2)}) months'),
'Meets 1-3-6', glue('{round((nrow(elssp%>%filter(Meets136=="yes"))/(nrow(elssp %>% filter(Meets136=="yes"|Meets136=="no"))))*100,2)}%')
),
      caption = "Meets 1-3-6 table") %>%
  kable_styling(latex_options = c("hold_position")) 
```

```{r variables-list}
tribble(
  ~Variable, ~Scale, ~ Range, 
'Age', 'Continuous', glue('{min(elssp$Age)}-{round(max(elssp$Age), digits=0)} months (mean (SD): {round(mean(elssp$Age))} ({round(sd(elssp$Age),1)}))'),
'Age at Amplification', 'Continuous', glue('{min(elssp$AgeAmplification, na.rm=TRUE)}-{max(elssp$AgeAmplification, na.rm=TRUE)} months (mean (SD): {round(mean(elssp$AgeAmplification, na.rm=TRUE))} ({round(sd(elssp$AgeAmplification, na.rm=TRUE),1)}))'),
'Age at Diagnosis', 'Continuous', glue('{min(elssp$IdentificationOfHLMonths, na.rm=TRUE)}-{max(elssp$IdentificationOfHLMonths, na.rm=TRUE)} months (mean (SD): {round(mean(elssp$IdentificationOfHLMonths, na.rm=TRUE))} ({round(sd(elssp$IdentificationOfHLMonths, na.rm=TRUE),1)}))'),
'Age at Implantation', 'Continuous', glue('{min(elssp$AgeImplantation, na.rm=TRUE)}-{max(elssp$AgeImplantation, na.rm=TRUE)} months (mean (SD): {round(mean(elssp$AgeImplantation, na.rm=TRUE))} ({round(sd(elssp$AgeImplantation, na.rm=TRUE),1)}))'),
'Age at Intervention', 'Continuous', glue('{min(elssp$AgeStartedServices, na.rm=TRUE)}-{max(elssp$AgeStartedServices, na.rm=TRUE)} months (mean (SD): {round(mean(elssp$AgeStartedServices, na.rm=TRUE))} ({round(sd(elssp$AgeStartedServices, na.rm=TRUE),1)}))'),
'Amplification', 'Categorical', 'Hearing Aid / Cochlear Implant / None',
'Communication', 'Categorical', 'Spoken / Total Communication / Cued Speech',
'Degree Hearing Loss (worse ear)', 'Continuous', glue('{min(elssp$HLworse, na.rm=TRUE)}-{max(elssp$HLworse, na.rm=TRUE)} dB HL (mean (SD): {round(mean(elssp$HLworse, na.rm=TRUE))} ({round(sd(elssp$HLworse, na.rm=TRUE))}))'),
'Developmental Delay', 'Categorical', 'Yes / No',
'Gender', 'Categorical', 'Female / Male',
'Health Issues', 'Categorical', 'Yes / No',
'Language in Home', 'Categorical', 'English / Other',
'Laterality', 'Categorical', 'Unilateral / Bilateral',
'1-3-6', 'Categorical', 'Yes / No',
'Prematurity', 'Categorical', 'Full-term / Premature',
'Services Per Month', 'Continuous', glue('{min(elssp$ServicesReceivedPerMonth, na.rm=TRUE)}-{max(elssp$ServicesReceivedPerMonth, na.rm=TRUE)} services per month (mean (SD): {round(mean(elssp$ServicesReceivedPerMonth, na.rm=TRUE))} ({round(sd(elssp$ServicesReceivedPerMonth, na.rm=TRUE),1)}))'),
'Etiology', 'Categorical', 'Sensorineural / Conductive / Mixed',
'CDI - Words Produced', 'Continuous', glue('{min(elssp$ProductionCDI, na.rm=TRUE)}-{max(elssp$ProductionCDI, na.rm=TRUE)} words (mean (SD): {round(mean(elssp$ProductionCDI, na.rm=TRUE))} ({round(sd(elssp$ProductionCDI, na.rm=TRUE),1)}))'),
) %>% kable(caption = "Variables table", digits=2) %>%
  kable_styling(latex_options = "scale_down") 
```

# Results
In the first section, we explore relationships among child demographic, audiological, and clinical variables. In the second section, we examine the influence of these factors on vocabulary development. In the third section, we describe the implementation of the EHDI 1-3-6 guidelines and predictors of early diagnosis and intervention. All analyses were conducted in R. All code is available on Github. 

## Part I: Interactions Among Variables

```{r graphs-for-eb}
ggplot((elssp %>% filter(Amplification != 'NA' & HLworse_cat != 'NA'))) +
  geom_bar(aes(x = HLworse_cat, fill = Amplification), position="dodge") +
  theme_classic()

ggplot((elssp %>% filter(Etiology != 'NA' & Laterality != 'NA' & Amplification != 'NA'))) +
  geom_bar(aes(x = Laterality, fill = Amplification), position="stack") +
  theme_classic() +
  facet_wrap(. ~ Etiology)
```

Before we test how these variables may be related to vocabulary, we describe their relationships to each other. As would be expected, many health, audiological, and clinical characteristics are not distributed randomly across this sample of children. To quantify this statistically, we used bonferroni-corrected chi-square tests between each of our variables (gender (male/female), laterality (bi-/uni-lateral hearing loss), health issues (yes/no), developmental delays (yes/no), prematurity (yes/no), language background (English/non-English), 1-3-6 (yes/no), degree of hearing loss (mild, moderate, severe/profound as defined above), etiology (sensorineural/ conductive), services received per month (binned into 0-2, 3-6, and >7 - to create maximally evenly sized bins), communication (spoken/total communication) and amplification (hearing aids/cochlear implants/none)). Because the chi-square statistic assumes n > 5 is *expected* in the majority of the cells for each test (preferably $\geq$ 80% @mchugh2013), we excluded mixed hearing loss (n=`r nrow(elssp %>% filter(Etiology=='Mixed'))`) and cued speech (n=`r nrow(elssp %>% filter(Communication=='cued speech'))`) from this section of the analysis. Strictly speaking, some of these variables are not expected to be randomly distributed relative to each other (e.g., prematurity and health issues; degree and amplification), but quantifying the differences via chi square using a conservative significance threshold lets us highlight the strongest relationships within this dataset.

Given that we ran `r nrow(chi_sq_all)` Chi-square tests, Bonferroni-corrected alpha for this set of analyses was p < 0.0007. Of these `r nrow(chi_sq_all)` combinations of variables, p < .05 for `r nrow(chi_sq_all %>% filter(p.value<.05))`, and `r nrow(chi_sq_all %>% filter(p.value<.000757))` survived Bonferroni correction. We are only discussing the latter below, but the full set of results can be found in figure \@ref(fig:relationships-plot).

```{r relationships-plot, fig.width=12, fig.height=9.5}
ggplot(chi_sq_all, aes(y=Var2, x=Var1)) +
  geom_tile(aes(fill = sig)) + theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust=0.90,vjust=0.2), 
        axis.text = element_text(size=16),
        legend.text = element_text(size=16)) +
  scale_fill_brewer(palette = "RdPu", direction=-1) +
  geom_text(aes(label=ifelse(sig=='survives bonferroni correction', round(eff.size,2), '')), 
            size=5) +
  xlab(NULL) + ylab(NULL) +
  labs(caption = "Results of chi-square tests between variables.\nFor tests that survived bonferroni correction,\neffect size (Cramer's V) is given.")
```

As expected, we found that health issues, developmental delays, and prematurity were highly interrelated in our sample, such that children born premature were more likely to also experience health issues `r chisq_output("IsPremature","HealthIssues")` and developmental delays `r chisq_output("IsPremature","DevelopmentalConcerns")`, and children with developmental delays were more likely to also experience health issues `r chisq_output("DevelopmentalConcerns","HealthIssues")`. Children with developmental delays received more services per month than typically-developing children `r chisq_output("DevelopmentalConcerns", "SPM_cat")` and were more likely to use total communication `r chisq_output("DevelopmentalConcerns","Communication")`. Likewise, children who used total communication received more services per month than children using spoken language `r chisq_output("Communication", "SPM_cat")`.

We also confirmed expected relationships among many of the audiological characteristics. There was a significant relationship between laterality and etiology `r chisq_output("Laterality","Etiology")`, such that children with conductive hearing loss were more likely to have unilateral hearing loss, children with sensorineural hearing loss were more likely to have a bilateral loss\footnote{All children with mixed hearing loss (n=`r nrow(elssp %>% filter(Etiology=='Mixed'))`) had bilateral hearing loss.}. Chi-square tests showed that laterality `r chisq_output("Laterality","Amplification")` and degree of hearing loss `r chisq_output("HLworse_cat", "Amplification")` were related to amplification in our sample. Children with bilateral hearing loss were more likely than children with unilateral hearing loss to use a hearing aid or cochlear implant; no child with unilateral hearing loss used a cochlear implant, and many children with unilateral hearing loss used no amplification. Regarding degree, children with severe-profound hearing loss were more likely to use a cochlear implant than children with less severe hearing loss (i.e., mild or moderate).

## Part II: Influence on vocabulary

We next turn to the relationship between each of these variables and children’s productive vocabulary, as measured by the CDI. Figure \@ref(fig:wg-ws-rainbow-plot) shows the vocabulary scores of children in our samples relative to norms for hearing children. Descriptively, we found widespread vocabulary delays on both Words and Gestures and Words and Sentences, with the majority of DHH children testing around or below the 25th percentile for hearing children (based on WordBank norms; @frank2017). 

As noted above, the CDI is composed of two instruments, which differ in number of questions (i.e. the maximum vocabulary score is 398 on Words and Gestures and 680 on Words and Sentences; 428 and 680 respectively for Spanish language CDI). To take this into account, rather than using the raw number of words produced as our outcome variable, we use the difference (in months) between the child’s chronological age and their predicted age based on their vocabulary from the WordBank norms (@frank2017). We call this derived variable **vocabulary delay**. 

More specifically, to compute a child's predicted age from their vocabulary score, we used the 50th percentile for productive vocabulary from Wordbank data from 8586 typically-developing infants [@frank2017] to create binary logistic growth curves separately for the Words and Gestures and Words and Sentences versions of the CDI for  American English and Mexican Spanish. For each child, we took the number of words they produced divided by the number of words on the instrument, to give us the proportion of words produced. We used this proportion in an inverse prediction from the binary logistic regression curves to generate a predicted age. That is, for each possible CDI score, the growth curve provided the age that score would be achieved for the 50th percentile trajectory. Finally, we subtracted the predicted age from each child's chronological age to calculate their vocabulary delay.
<!-- ec: leaving until we figure out what to do with growth curve -->
<!-- ec: come back to #s and change after we figure out which data to use. how to make it populate automatically? -->


```{r wg-ws-rainbow-plot}

WG_eng_curves <- ggplot()  + 
  geom_line(data = WG_elssp_eng_gcurve, aes(x=predict_ages, y=scores),colour='black', alpha= .5) + 
  theme_classic() + 
  ylab("CDI Score (WG English)") + 
  xlab("Age in Months") + 
  coord_cartesian(xlim=c(0,40)) + 
  geom_point(data = subset(WG_elssp_eng, !is.na(ProductionCDI)), aes(x= Age, y= ProductionCDI))+
  theme(legend.position = "none")

WS_eng_curves <- ggplot()  + 
  geom_line(data = WS_elssp_eng_gcurve, aes(x=predict_ages, y=scores),colour='black', alpha= .5) + 
  theme_classic() + 
  ylab("CDI Score (WS English)") + 
  xlab("Age in Months") + 
  coord_cartesian(xlim=c(0,40)) + 
  geom_point(data = subset(WS_elssp_eng, !is.na(ProductionCDI)), aes(x= Age, y= ProductionCDI)) +
  theme(legend.position = "none")

WG_span_curves <- ggplot()  + 
  geom_line(data = WG_elssp_span_gcurve, aes(x=predict_ages, y=scores),colour='black', alpha= .5) + 
  theme_classic() + 
  ylab("CDI Score (WG Spanish)") + 
  xlab("Age in Months") + 
  coord_cartesian(xlim=c(0,40)) + 
  geom_point(data = subset(WG_elssp_span, !is.na(ProductionCDI)), aes(x= Age, y= ProductionCDI))+
  theme(legend.position = "none")

WS_span_curves <- ggplot()  + 
  geom_line(data = WS_elssp_span_gcurve, aes(x=predict_ages, y=scores),colour='black', alpha= .5) + 
  theme_classic() + 
  ylab("CDI Score (WS Spanish)") + 
  xlab("Age in Months") + 
  coord_cartesian(xlim=c(0,40)) + 
  geom_point(data = subset(WS_elssp_span, !is.na(ProductionCDI)), aes(x= Age, y= ProductionCDI)) +
  theme(legend.position = "none")

cowplot::plot_grid(WG_eng_curves, WS_eng_curves) +
  labs(caption = "Growth curve from Wordbank American English 50th percentile data. Black triangles show vocabulary scores of individual DHH children.")
cowplot::plot_grid(WG_span_curves, WS_span_curves) +
  labs(caption = "Growth curve from Wordbank Spanish (Mexican) 50th percentile data. Black triangles show vocabulary scores of individual DHH children.")

```

To look at the relationship between our predictor variables and CDI scores, we next conducted multiple linear regression, using vocabulary delay as our outcome variable. \footnote{We excluded the adopted child from this section of the analysis due to concerns about comparing her score to the American English CDI norms.}

Our full regression model included all variables: 

Vocabulary Delay ~ Gender + Developmental Delay + Health Issues + Prematurity + Laterality + Degree + Amplification + Communication + Meets 1-3-6 + ServicesReceivedPerMonth + LanguageBackground. 



```{r delay-AIC, include=F}

full_model_delay <- lm(diff_age_from_expected ~ Gender + DevelopmentalConcerns + IsPremature + HealthIssues + Laterality + Amplification + Communication + Meets136 + PrimaryLanguage + HLworse +Age, data = (elssp_curves %>% filter(PrimaryLanguage!='Hindi')))
summary(full_model_delay)
# stepAIC(full_model_delay)
#need to filter out cases where degree or meets136 are blank. AIC won't run with missing values in those columns

full_model_delay_filtered <- lm(diff_age_from_expected ~ Gender + DevelopmentalConcerns + IsPremature + HealthIssues + Laterality + Amplification + Communication + Meets136 + PrimaryLanguage + HLworse +Age, data = (elssp_curves %>% filter(PrimaryLanguage!='Hindi' & Meets136!='' & HLworse!='')))
stepAIC(full_model_delay_filtered)
#removed HLworse (degree) and tried again because I had to remove participants to run AIC with it in, and it didn't improve model fit anyway

full_model_delay2 <- lm(diff_age_from_expected ~ Gender + DevelopmentalConcerns + IsPremature + HealthIssues + Laterality + Amplification + Communication + Meets136 + PrimaryLanguage +Age, data = (elssp_curves %>% filter(PrimaryLanguage!='Hindi' & Meets136!='')))
stepAIC(full_model_delay2)

AIC_model_delay <- lm(formula = diff_age_from_expected ~ Laterality + Amplification + Age, data = (elssp_curves %>% filter(PrimaryLanguage != 
    "Hindi")))
summary(AIC_model_delay)

vif(AIC_model_delay)
```


```{r full-model-betas}
dwplot(full_model_delay) +
  geom_vline(xintercept=0, linetype = 'dashed') + 
  theme_classic() + 
  theme(legend.position="none")
options(repr.plot.width=6, repr.plot.height=4) 
```

This model accounted for significant variance in vocabulary delay (adjusted-R^2^ = `r summary(AIC_model_delay)$adj.r.squared`, *p* = `r lm_pvalue(AIC_model_delay)`; see figure XXX). We next performed stepwise model comparison using stepAIC (MASS) to pare down the model. This process selected only the predictors which incrementally improved model fit, measured by Akaike’s Information Criterion (AIC), which considers goodness of fit and model complexity (penalizing models with many predictors). Based on this iterative process, we removed several predictors from the model. 
<!-- eb2ec: 
also: clarify when you describe language background:for families where there was a spanish cdi was there ALSO an english cdi? ) -->

Thus, our final model was: Vocabulary Delay ~ Age + Laterality + Amplification. This model accounted for significant variance in children's vocabulary delay (adjusted-R^2^ = `r summary(AIC_model_delay)$adj.r.squared`, *p* = `r lm_pvalue(AIC_model_delay)`). We found significant main effects for Age (`r beta_output(AIC_model_delay, 5)`), Amplification (Hearing Aid: `r beta_output(AIC_model_delay, 3)`; No amplification: `r beta_output(AIC_model_delay, 4)`), and Laterality (`r beta_output(AIC_model_delay, 2)`), such that older age, no amplification, and bilateral hearing loss predicted greater delays. Although we showed in Part I that relationships exist among several of these variables (e.g., laterality and amplification), the vif (variance inflation factor) for our model revealed low levels of collinearity among our predictors (all VIF < 1.5; see table \@ref(tab:delay-betas); @james2013).

<!-- eb2ec: something is not right in the preceding para, see rendered text -->

```{r delay-betas, results="asis"}
dw_plot(AIC_model_delay) %>%
  relabel_predictors(DevelopmentalConcernsyes = "Developmental Delay (yes)",
Meets136yes= "1-3-6 (yes)",
PrimaryLanguageSpanish = "Language Background (Spanish)")  + 
  theme_classic() + 
  geom_vline(xintercept = 0) + 
  theme(legend.position = "none")
# eb: also this graph can be much smaller and still fine

AIC_model_delay %>% 
  tidy() %>% 
  kable() %>%
  kable_styling()

# kable((vif(AIC_model_delay))[,1:2]) %>% 
#   kable_styling() this doesn't work anymore. need to troubleshoot
```

## Part III: Meets136 success
```{r 136-summary-stats}
perc_136 <- (nrow(elssp%>%filter(Meets136=='yes'))/(nrow(elssp%>%filter(Meets136=='yes'|Meets136=='no'))))*100
perc_136_TD <- (nrow(elssp%>%filter(anycomorbid=='no')%>%filter(Meets136=='yes'))/(nrow(elssp%>%filter(anycomorbid=='no')%>%filter(Meets136=='yes'|Meets136=='no'))))*100
perc_136_DA <- (nrow(elssp%>%filter(anycomorbid=='yes')%>%filter(Meets136=='yes'))/(nrow(elssp%>%filter(anycomorbid=='yes')%>%filter(Meets136=='yes'|Meets136=='no'))))*100
```

Perhaps of greatest importance to clinicians and policymakers is the implementation and effect of existing policies. Although whether a child met 1-3-6 guidelines was not significant in our final model predicting vocabulary delay, its demonstrated importance for practice and policy (e.g., XXX) merit further discussion. To this end, we looked at the ages at which children received diagnosis and intervention, and how this mapped onto the 1-3-6 guidelines.  In this section, we provide a brief description of the implementation of 1-3-6 in our sample, examine its effect on vocabulary delay, and describe the results of exploratory linear regression models for age at diagnosis and age at intervention. 
<!-- eb: do you mean that 1-3-6 wasn't in the final model? bc that doesn't mean it didnt' significantly predict vocab delay. is there a reasonable correlation between 1-3-6 & vocab delay? 
ec: no :( 
okay, this is maybe not true anymore. depends what we do with the growth curves -->

Overall, `r round(perc_136)`% of our sample met 1-3-6 guidelines for early diagnosis and intervention (see \@ref(tab:CDIinfo)). Among the children for which screening information was available (n=`r nrow(elssp %>% filter(NBHS=='Y'))`), 100% were screened at birth or during NICU stay. `r round((nrow(elssp%>%filter(meets13=="yes"))/(nrow(elssp %>% filter(meets13=="yes"|meets13=="no"))))*100)`% of children received diagnosis by 3 months of age, and `r round((nrow(elssp%>%filter(meets6=="yes"))/(nrow(elssp %>% filter(meets6=="yes"|meets6=="no"))))*100)`% began early intervention by 6 months of age. Among children with comorbidities, `r perc_136_DA`% met 1-3-6 guidelines, compared to `r perc_136_TD`% of children without comorbidities. Figure \@ref(fig:meets136-timeline) shows the age at first diagnosis, intervention, amplification, and implantation for each child in our sample.


```{r meets136-timeline}
event_summary <- function(x) {
   m <- mean(x)
   ymin <- m-sd(x)
   ymax <- m+sd(x)
   return(c(y=m,ymin=ymin,ymax=ymax))
}

elssp %>% 
  dplyr::select(VIHI_ID, IdentificationOfHLMonths, AgeStartedServices, AgeAmplification, AgeImplantation) %>%
  rename('Age at Diagnosis'=IdentificationOfHLMonths,
         'Age at Intervention' = AgeStartedServices,
         'Age at Implantation'= AgeImplantation,
         'Age at Amplification' = AgeAmplification) %>%
  pivot_longer(cols = 'Age at Diagnosis':'Age at Implantation', names_to = "event", values_to = "age") %>%
  ggplot(aes(x = event, y = age, fill = event, color = event)) +
  geom_violin(alpha = .4) +
  geom_jitter() + 
  coord_flip() +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = '', y ='Age (months)') +
stat_summary(fun.data=event_summary, color = "black", geom="pointrange")
```



```{r delay-violins}
t_136 <- t.test(elssp_curves$diff_age_from_expected ~ elssp_curves$Meets136)

gg_SE <- function(x) {
   m <- mean(x)
   ymin <- m-se(x)
   ymax <- m+se(x)
   return(c(y=m,ymin=ymin,ymax=ymax))
}

ggplot(data=(elssp_curves %>% filter(Meets136!='')), aes(x=Meets136, y=diff_age_from_expected, fill=Meets136, color=Meets136)) +
  geom_violin(alpha=.4)+
  theme_classic() +
  geom_jitter() +
  stat_summary(fun.data=gg_SE, mult=1, 
                 geom="pointrange", color="black") +
  xlab("1-3-6") +
  ylab("Vocabulary Delay (months)") + 
  coord_flip()

```

Although 1-3-6 was not a significant main effect in our model of vocabulary delay, an independent samples t-test shows that children who did not meet 1-3-6 guidelines had significantly larger vocabulary delays than children who met 1-3-6 guidelines (*t*(`r t_136$parameter`)=`r t_136$statistic`, *p*=`r t_136$p.value`; see Figure \@ref(fig:delay-violins)).


```{r lm-stats, include=F}
# i want to look through the residuals plots with you
stepAIC(lm(IdentificationOfHLMonths ~ Gender + 
             Etiology +
             HLworse + 
             Laterality + 
             IsPremature + 
             HealthIssues + 
             DevelopmentalConcerns + 
             PrimaryLanguage + 
             VisionLoss + 
             ANSD, 
           (data=elssp %>% 
              filter(HLworse!="NA" & Etiology!='' & PrimaryLanguage!='Hindi'))))
#Health issues , primary language
diagnosis_aic <- lm(formula = IdentificationOfHLMonths ~ HealthIssues + Monolingual_English + Laterality, data = elssp)


stepAIC(lm(AgeStartedServices ~ Gender + 
             Etiology +
             HLworse + 
             Laterality + 
             HealthIssues + 
             IsPremature + 
             DevelopmentalConcerns + 
             Monolingual_English + 
             Communication + 
             VisionLoss + 
             ANSD + 
             IdentificationOfHLMonths,
           (data=elssp %>% filter(HLworse!="NA"& Etiology!='')))) 
services_aic <- lm(formula = AgeStartedServices ~ HLworse + IsPremature + 
                     IdentificationOfHLMonths + Monolingual_English, data = elssp)



difference_aic <- lm(formula = (AgeStartedServices - IdentificationOfHLMonths) ~ 
    HLworse + IsPremature + Monolingual_English + IdentificationOfHLMonths, 
    data = elssp)
```

To better understand implementation of 1-3-6 guidelines, we zoomed in on diagnosis and intervention. We conducted two linear regressions, one for age at diagnosis and one for age at intervention. For each model, we started with the set of predictors that would have been present prior to or during diagnosis or intervention, respectively. We then pared down each model using stepwise regression by AIC (MASS package), using the process described above in Part II of Results. 

For age at diagnosis, we included the set of child-specific factors that would be relevant *before* diagnosis of hearing loss (e.g., we excluded amplification type because a child would not receive a hearing aid or cochlear implant prior to being diagnosed with hearing loss.) We began with: gender, degree, developmental delay, health issues, prematurity, laterality, language background, and etiology. 

The best fit model was: Age at Diagnosis ~ Health Issues + Language Background + Laterality, with significant main effects of Health Issues and Language Background. This model accounted for `r (summary(diagnosis_aic)$r.squared)*100`% of the variance in age at diagnosis (p=`r lm_pvalue(diagnosis_aic)`). Average age at dianosis for English-speaking children with bilateral hearing loss and no health issues was `r summary(diagnosis_aic)$coefficients[1,1] + summary(diagnosis_aic)$coefficients[3,1]` months. Relative to English-speaking families, children from Spanish-speaking families were diagnosed `r abs(summary(diagnosis_aic)$coefficients[3,1])` months later (p=`r abs(summary(diagnosis_aic)$coefficients[3,4])`). Children with health issues were diagnosed `r summary(diagnosis_aic)$coefficients[2,1])` months later than children without health issues (p=`r abs(summary(diagnosis_aic)$coefficients[2,4])`).

<!-- eb2ec: you should tell the reader what the best fit model is before you tell us about individually significant  predictors, but actaully it looks like you're describing all predictors in the final model even if they're not significant on their own, which is fine,  just be clearer about this.
you may also want to add full regression tables? -->
<!-- eb2ec: r^2 and % variance accounted for are the same thing so consolidate those pieces of info in the preceiding two sentences -->

```{r diagnosis-betas, results="asis"}
dwplot(diagnosis_aic) %>%
  relabel_predictors(HealthIssuesyes = "Health Issues (yes)",
Monolingual_Englishyes = "Language Background (English)",
LateralityUnilateral = "Laterality (Unilateral)") + 
  geom_vline(xintercept=0, linetype = 'dashed') + 
  theme_classic()+ 
  theme(legend.position="none")
options(repr.plot.width=6)
#caption: 

diagnosis_aic %>% 
  tidy() %>% 
  kable() %>%
  kable_styling()

kable(rownames_to_column(data.frame(vif(diagnosis_aic))) %>% 
  mutate(Df = 1) %>%
  rename(VIF = 'vif.diagnosis_aic.', ' ' = rowname)) %>%
  kable_styling()
```


We repeated this model selection process for age at intervention. In addition to the variables used to fit the intervention model, we included age at diagnosis. The best fit model was Age Intervention ~ prematurity + degree + age at diagnosis + language background (R^2^=`r summary(services_aic)$r.squared` , p=`r lm_pvalue(services_aic)`), with significant main effects of degree and age at diagnosis. Less severe hearing loss `r beta_output(services_aic, 2)` and later diagnosis `r beta_output(services_aic, 4)`, and coming from a non-English-speaking household `r beta_output(services_aic, 5)` predicted later intervention. Prematurity and language background were not significant, but their inclusion improved model fit.

```{r intervention-betas}
services_aic_df <- broom::tidy(services_aic) 

dwplot(services_aic)  %>%
relabel_predictors(HLworse = "Degree of Hearing Loss",
IsPrematureyes = "Premature (yes)",
IdentificationOfHLMonths = "Age at Diagnosis",
Monolingual_Englishyes = "Language Background (English)") +
  geom_vline(xintercept=0, linetype = 'dashed') + 
  theme_classic() + 
  theme(legend.position="none")
options(repr.plot.width=6, repr.plot.height=4) 
#caption:  

services_aic %>% 
  tidy() %>%
  kable() %>% 
  kable_styling()

kable(rownames_to_column(data.frame(vif(services_aic))) %>% 
  mutate(Df = 1) %>%
  rename(VIF = 'vif.services_aic.', ' ' = rowname)) %>%
  kable_styling()
```
# Discussion

In this study, we examined the demographic, audiological, and clinical characteristics of 100 young DHH children in North Carolina. We documented the distribution of these characteristics and explored the relationships between these variables, vocabulary, diagnosis, and intervention. This analysis was exploratory and descriptive, and the results should be interpreted accordingly. 

## How are child-level variables intertwined?

In our sample, we found significant overlap among demographic, audiological, and clinical variables. Prematurity, health issues, and developmental delay frequently co-occurred, such that there was a moderate relationship between each of these variables (cramer’s V = 0.38 - 0.52, p < .0007). Children with one of these conditions (prematurity, developmental delay, health issues) were more likely to have any other condition. This is not surprising. Many conditions that cause developmental delays have a high incidence of health issues (e.g., heart problems in Down Syndrome; vomiting and seizures with hydrocephalus), and it is well documented that there is a higher incidence of developmental delay in preterm infants [@pierrat2017; @aarnoudse-moens2009]. Children born premature, especially those born extremely premature, are at increased risk for a number of health issues at birth [@robertson2009; @costeloe2012; @york2002] and throughout the lifespan [@luu2016]. 

Each of these conditions may affect language and development in different ways. The literature points to increased risk of language delay for children with developmental delays [@chapman1997; @kristoffersen2008; @weismer2010] and children born premature [@foster-cohen2010], with differential effects based on the nature of the child's diagnosis [@cupples2018; @cupples2014]. Together, these risks may interact and multiply. In our sample, we also had a large range of health conditions (76 unique conditions in our sample of 100 children; see \@ref(tab:comorbid-info) and Appendix XXX for more detailed information about comorbidities), and it appears probable that those conditions would vary in whether and how they influence vocabulary growth. 

We found that children with developmental delays (e.g., Down syndrome) were much more likely to use a total communication approach than typically-developing DHH children. Assignment to “spoken language” and “total communication” groups was not randomly distributed, with use of total communication appearing to follow children already at greater risk for verbal delays. Such a pattern is in line with clinical use of manual communication approaches for young children with disabilities [e.g., @branson2009]. This should temper the interpretation of correlational studies finding links between total communication and language delays [e.g., @geers2017].

Additionally, in our sample, children with developmental delays were considerably more likely to receive >10 services per month, perhaps accounting for increased need (or increased perceived need). The services per month variable also includes occupational therapy, physical therapy, which typically-developing DHH children may be unlikely to receive. Likewise, children who used total communication were more likely to receive frequent services.

We also found relationships among many of our audiological variables. In particular, etiology and laterality were related, such that conductive hearing loss was more likely unilateral, and sensorineural hearing loss was more likely bilateral. There were only seven cases of mixed hearing loss, and all were bilateral. One possible explanation is that certain underlying causes of conductive hearing loss (e.g., aural atresia, impacted cerumen, trauma to the tympanic membrane) may be more likely to affect one ear than two.

Amplification devices were more common for children with less hearing (i.e., children with bilateral hearing loss and children with moderate to profound hearing loss). This may be due to the assumption that a hearing aid or cochlear implant will not benefit children with minimal hearing loss [@updike1994], although several studies have found benefits in speech perception and quality of life for amplification for unilateral hearing loss  [@hassepass2013; @priwin2007; @briggs2011; @winiger2016] and spoken language vocabulary and grammar for mild hearing loss [@walker2015].

These results are not necessarily surprising, given causal links among some of the variables (e.g., increased health issues in children born premature). Nevertheless, it should caution us to think critically about how we construct samples for controlled lab experiments. During study design: how likely is it to collect a desired sample of (e.g.) 32 typically-developing pediatric cochlear implant with bilateral, severe-to-profound hearing loss, given that such a subsample may only represent `r nrow(elssp %>% filter(Laterality=='Bilateral' & HLworse>70 & Amplification=='CI' & anycomorbid=='no'))`% of the DHH population? During interpretation of the results: how might the findings generalize to the rest of the DHH population given the constraints of the study?

## Predicting vocabulary outcomes

`r ((nrow(elssp_curves %>% filter(diff_age_from_expected > 0))) / (nrow(elssp_curves %>% filter(diff_age_from_expected != '')))) * 100`% of DHH children in our sample fell below the 50th percentile for spoken vocabulary. Of the `r ((nrow(elssp_curves %>% filter(diff_age_from_expected <= 0))) / (nrow(elssp_curves %>% filter(diff_age_from_expected != '')))) * 100`% who were at or above the 50th percentile, `r nrow(elssp_curves %>% filter(diff_age_from_expected < 0 & Age > 9))`/`r nrow(elssp_curves %>% filter(diff_age_from_expected < 0))` were 8-to-9-month olds who were not yet producing any words (as expected for 8-to-9-month-olds). To have such a strong of majority DHH children below the 50th percentile for vocabulary development indicates that this group is not yet well-equipped to acquire spoken language. This disadvantage can have lasting consequences in the lives of DHH children [@kyle2010; @qi2012; @karchmer2003]. 

We had hypothesized that male gender, more severe degree of hearing loss,  bilateral hearing loss, no amplification use, prematurity, and presence of additional disabilities would predict larger spoken vocabulary delay. However, in our analysis, we found that age, amplification, and laterality of hearing loss best explained the variance in vocabulary outcomes. We found a significant main effect of age. This could be explained in part by constraints on the variable. A 9-month-old cannot be 12 months delayed, but a 24-month-old can. However, if DHH children were learning words at the same rate (albeit delayed) as hearing children, we would expect delay to remain constant across time. Instead, we see that the spoken vocabulary delay widens with age, indicating that the rate of spoken vocabulary acquisition is slower for DHH children. The result is a population increasingly behind on spoken language milestones, and because none of the children use sign language, on language development more broadly. 

An alternative explanation is that the age effect is related to age at intervention. We analyzed the first evaluation available from each participant, which was often (but not always) their first evaluation with the early intervention program. Perhaps children who were older at first evaluation were more likely to have a wider delay. Exploring the longitudinal administrations would help us clarify whether age effect reflects an artifact of our sampling or an increasing delay with age.

In our sample, children without amplification were `r abs(summary(diagnosis_aic)$coefficients[4,1])` months more delayed than children with a cochlear implant. Vocabulary delay was not significantly different between children with hearing aids and children with cochlear implants. This increased delay for children without amplification could reflect both increased spoken language audibility and/or a difference in demographic characteristics of amplification candidates / non-candidates. Of the typically-developing children in our sample, `r nrow(elssp %>% filter(Amplification=="CI" & DevelopmentalConcerns=="no")) / nrow(elssp %>% filter(DevelopmentalConcerns=="no")) * 100`% received cochlear implants, `r nrow(elssp %>% filter(Amplification=="HA" & DevelopmentalConcerns=="no")) / nrow(elssp %>% filter(DevelopmentalConcerns=="no")) * 100`% of children received hearing aids, and `r nrow(elssp %>% filter(Amplification=="none" & DevelopmentalConcerns=="no")) / nrow(elssp %>% filter(DevelopmentalConcerns=="no")) * 100`% received no amplification. Meanwhile, for children with developmental concerns, `r nrow(elssp %>% filter(Amplification=="CI" & DevelopmentalConcerns=="yes"))`% received cochlear implants, `r nrow(elssp %>% filter(Amplification=="HA" & DevelopmentalConcerns=="yes")) / nrow(elssp %>% filter(DevelopmentalConcerns=="yes")) * 100`% received hearing aids, and `r nrow(elssp %>% filter(Amplification=="none" & DevelopmentalConcerns=="yes")) / nrow(elssp %>% filter(DevelopmentalConcerns=="yes")) * 100`% received no amplification (p<.05, but does not survive Bonferroni correction; see Figure \@ref(fig:relationships-plot)).

In our model, children with bilateral hearing loss had an `r abs(summary(diagnosis_aic)$coefficients[2,1])` months greater delay than children with unilateral hearing loss. While few studies directly assess language development differences between unilateral and bilateral hearing loss, we know that both bilateral and unilateral hearing loss have detrimental effects on spoken language development [XXX]. Our model results suggest that children with bilateral hearing loss are at a quantifiable disadvantage over children with unilateral hearing loss. This falls in line with research that finds better spoken language outcomes for individuals with more residual hearing [@lieu2013;@anne2017; @vohr2008; @tomblin2015]. However even children with unilateral hearing loss, who have one ear with normal hearing, experience notable delays both in the literature [@vila2015; @lieu2013; @lieu2004; @lieu2012; @kiese-himmel2002] and in our sample (Mean delay unilateral = `r mean((elssp_curves%>% filter(Laterality=='Unilateral'))$diff_age_from_expected)`). 

## Predicting early diagnosis and intervention

Our findings from North Carolina parallel AAP’s national findings: approximately 70% of children in our sample were diagnosed by 3 months, and only about 40% began services by 6 months. Only 36% of children met the EHDI guidelines, despite ample evidence suggesting early diagnosis and intervention improve language outcomes [@yoshinaga-itano2018; @yoshinaga-itano1998; @white1987; @robinshaw1995; @apuzzo1995; @kennedy2006; @vohr2008; @watkin2007; @ching2013; @holzinger2011; @vohr2011]. 

In support of this policy, Ching and colleagues [@ching2013] find a pronounced effect of early diagnosis. In this study, which was conducted in Australia, researchers compared groups of children in regions that had implemented universal newborn hearing screenings to children in regions that had not yet implemented UNHS. Children were otherwise similar in diagnostic, demographic, and intervention characteristics. The UNHS group had higher global language scores. By contrast, in our sample, by dint of accepting all children receiving early intervention services in one state, we were able to document naturally occurring variance in who received on-time diagnosis and intervention. We found that some of the variance in age at diagnosis and intervention could be explained by children’s demographic and audiological characteristics. We hypothesized that children with less residual hearing (i.e., bilateral, more severe) and no co-occurring conditions would be earlier diagnosed and earlier to begin language services, and that earlier diagnosis would predict earlier intervention.

### Diagnosis 

In the case of diagnosis, presence of health issues, primary language in the home, and laterality of hearing loss accounted for `r summary(diagnosis_aic)$r.squared`% of variance in age at diagnosis. Having diagnosed health issues and non-English language background predicted later diagnosis; laterality was not significant. 

Children with health issues were diagnosed `r beta_output(diagnosis_aic, 2)` months later than infants without health issues. One possible explanation is that the health issues caused acquired hearing loss that wouldn’t be detected by the NBHS, thus delaying identification of hearing loss. Of the 76 unique health issues experienced by children in the sample, only 9 conditions might cause acquired hearing loss (i.e., meningitis, sepsis, jaundice, seizures, hydrocephalus, MRSA, anemia, frequent fevers, cytomegalovirus, affecting `r nrow(comorbid %>% filter(pos_acqHL == 1))` out of `r nrow(elssp %>% filter(HealthIssues=='yes'))` children with health issues in our sample. Another possible explanation is that the health issues required more pressing medical attention than the possible hearing loss, and that families and medical providers had to prioritize treatment for the health issue (e.g., surgery for congenital heart defect) over diagnostic audiology services.

Infants from Spanish-speaking families were diagnosed `r abs(summary(services_aic)$coefficients[3,1])` months later than infants from English-speaking families. This may be due to cultural differences in attitudes towards deafness [@caballero2018; @steinberg1997, @steinberg2003; @rodriguez2020] or it may result from a lack of linguistically accessible and culturally appropriate audiology services. Only 5.6% of American audiologists identify as a bilingual service provider [@asha2019], and services from a monolingual provider may be insufficient. @caballero2017 found that Hispanic-American parents of DHH children want more concrete resources, comprehensive information, and emotional support from their audiologist. In a nationwide survey, the majority of audiologists reported using ad-hoc interpreters (e.g., client’s family) to overcome language barriers. Survey respondents reported that language barriers presented a major challenge in working with Spanish-speaking families, specifically in obtaining the child’s case history and providing recommendations for follow-up services [@abreu2011]. 

### Intervention
In line with our prediction, more severe hearing loss predicted earlier intervention, such that for every additional 10 dB HL, predicted age at diagnosis was `r abs((summary(services_aic)$coefficients[2,1])*10*4.348)` weeks earlier. This parallels findings by @harrison2003 in which severe-to-profound hearing loss was diagnosed 2-5 months later than mild-to-moderate hearing loss. This could be related to diagnosis; certain methods of hearing screening (i.e., Auditory Brainstem Responses) can miss milder hearing loss [@johnson2005], and parents may not suspect hearing loss in children who respond to loud noises. Alternatively, parents and clinicians may adopt a wait-and-see approach to intervention for children with some residual hearing. Nevertheless, mild-to-moderate hearing loss is associated with language delays and academic challenges [@delage2007; @blair1985], which early intervention may offset.

Age at start of services was also associated with age at diagnosis. This result was in line with our hypotheses. For every month diagnosis was delayed, intervention was delayed by `r (summary(services_aic)$coefficients[4,1])*4.348` weeks. Early diagnosis puts children in the pipeline towards intervention earlier. @ching2013 found that age at intervention predicted better outcomes for DHH children, above and beyond age at diagnosis. Unsurprisingly however, these two variables are related, such that we cannot hope to achieve early intervention goals without ensuring children receive timely diagnosis. 

This sample is composed of children receiving birth-to-3 services. An estimated 67% of children with hearing loss enroll in early intervention services [@cdc2018]. While this represents a tremendous step forward in prompt early intervention services, early intervention may not be early enough. Less than `r round((nrow(elssp%>%filter(meets6=="yes"))/(nrow(elssp %>% filter(meets6=="yes"|meets6=="no"))))*100)`% of our sample of children in early intervention meet the 6-month EHDI benchmark. Furthermore, critically, there exists a significant chunk of the population who aren’t included in this analysis and for whom we don’t have any data because they have not been enrolled in services by 36 months. The AAP estimates that almost 36% of infants who do not pass a newborn hearing screening are lost to follow-up. Assuming that the population of children in early intervention only represents two thirds of the population with hearing loss, our data suggest that the actual proportion of DHH children who receive intervention by the EHDI-recommended 6 months may be closer to `r round((nrow(elssp%>%filter(meets6=="yes"))/(nrow(elssp %>% filter(meets6=="yes"|meets6=="no"))*3/2))*100)`. These children may not receive clinical support until school-age or later. 

## Educational and Clinical Implications

Despite high rates of NBGS in North Carolina, and even relatively high rates of diagnosis by 3 months (66/100 children in our sample), most children in our sample did not meet the 1-3-6 guidelines. Based on our analyses, we have the following recommendations for increasing attainment of 1-3-6 guidelines:\begin{enumerate}

\item Frequent hearing screenings for children receiving medical or therapeutic care for health issues.
\item Service coordination for families balancing multiple co-occurring conditions.
\item Expansion of bilingual clinicians both in-person and teletherapy clinicians to provide therapy and service coordination to non-English-speaking families.
\item Provision and encouragement of early intervention services for children with mild to moderate hearing loss.

\end{enumerate}

Additionally, the vast majority of children in our sample experienced vocabulary delays (relative to hearing peers), and studies of spoken vocabulary development in older DHH children suggest that they may not catch up [@lund2016]. This should set clinicians and educators on high alert, due to the demonstrated importance of vocabulary skills in literacy [@staehr2008; @hemphill2008; @biemiller2003] and in education more broadly [e.g., @young2005; @monroe2002]. As early intervention predicts vocabulary outcomes in study after study [including this present study and e.g., @vohr2008; @watkin2007; @ching2013; @holzinger2011; @vohr2011; @ching2018], ensuring intervention by 6 months for all DHH children may be one way to address spoken vocabulary deficits. Prior to intervention or amplification, provision of accessible language input (i.e., sign language) may mitigate negative effects of auditory deprivation on language skills [@davidson2014; @hassanzadeh2012; @spellun2018]. 

We encourage sign language *at least* prior to mastery of spoken language, and when possible for the family, we encourage its continued use as a language resource. While some children in our sample exhibited vocabulary development at or even slightly above the 50th percentile for hearing peers (n=`r nrow(elssp_curves %>% filter(diff_age_from_expected <= 0))`), the vast majority of children fell below this threshold (n=`r nrow(elssp_curves %>% filter(diff_age_from_expected > 0))`). Even in the **best** case scenario for spoken language development (i.e., unilateral hearing loss, uses amplification, etc.), children in our sample did not excel with spoken language. Offering from-diagnosis access to structured linguistic input through sign language may help overcome this deficit. In recommending sign language, we follow the rationale set forth by @hall2019, summarized here: Spoken language outcomes for DHH children are variable and unpredictable [@ganek2012; @szagun2016], and even in optimal situations, many DHH children do not achieve age-appropriate spoken language outcomes [@geers2017; XXX]. Failing to achieve language proficiency (in any language) confers higher risk of disrupted cognitive, academic, and socioemotional development [@moeller2006; @schick2007; @preisler2002; @hrastinski2016; @dammeyer2010; @desselle1994; @kushalnagar2011; @amraei2017; @hall2017]. The available data do not suggest that sign language harms spoken language development [@park2013; @davidson2014], and in fact, some studies suggest that sign language *benefits* spoken language development [e.g., @hassanzadeh2012]. Providing early access to a natural sign language offers children the best chance of achieving language mastery, and use of sign language does not preclude learning spoken language.

# Limitations

These analyses were exploratory, and there were many possible analytic routes. In the interest of transparency, these data are available on our OSF page (XXX), and all of the code used to generate the statistics and figures from this article are available on Github (XXX). With that being said, our results largely replicated past studies [e.g., @ching2013] or clinical intuitions. Additionally, many children in our sample have multiple evaluations, so we have an opportunity to see whether these results hold up longitudinally. 

This sample is comprised only of children in North Carolina, and certain factors, such as diagnosis and early intervention practices, vary by country and by state [@nad]. However, based on Gallaudet's 2014 Annual Survey of Deaf and Hard-of-Hearing Children and the SEELS report of elementary and middle school aged DHH children [@blackorby2006; @gallaudetresearchinstitute2014], our sample largely resembles the national DHH population in terms of degree of hearing loss, percentage of children with additional disabilities, cochlear implant and hearing aid use, and gender. Descriptively, compared to the Gallaudet sample, more children in our sample spoke Spanish, and many fewer children in our sample used sign language. We would exercise caution in applying these results to regions where sign language access for DHH children is more common, such as Washington D.C. or Rochester, New York. A similar naturalistic study in those regions could help illuminate the effects of different clinical and demographic factors in a signing population

There were several variables reported to influence outcomes for DHH children that we did not have access to (e.g., SES, maternal education, parental hearing status). Measuring these variables may have provided additional predictive power. Furthermore, the considerable variability in the sample did not allow us to easily isolate effects of different characteristics. However, this variability is real-world variability, and as we demonstrated earlier, many of these variables co-occurred such that it may not make sense to isolate. Larger Ns, which are often difficult to achieve in research with DHH children, would help to tease apart different effects. As researchers continue to study influences on vocabulary in DHH children, a meta-analytic approach may be able to better estimate effects and effect sizes within the varied outcomes of this diverse population.

# Conclusion

Using a diverse sample of 100 children enrolled in early intervention, we provide a description of children's demographic and audiological characteristics, vocabulary outcomes, and clinical milestones. Many variables in this sample co-occurred disproportionately (e.g., amplification and degree of hearing loss; communication modality and developmental delay); this paper provides the first population-based documentation of this distribution. The vast majority of vocabulary scores in our sample were well below established norms for hearing infants. Significant predictors of vocabulary delays were: age, amplification, and laterality. Only `r round((nrow(elssp%>%filter(Meets136=="yes"))/(nrow(elssp %>% filter(Meets136=="yes"|Meets136=="no"))))*100)`% of children met 1-3-6 guidelines for early detection and intervention. English language background and having no co-occurring health issues significantly predicted earlier diagnosis. More severe hearing loss and earlier diagnosis significantly predicted earlier intervention.

# Acknowledgments 

\pagebreak
# References